Equations & R Code

### 2017-01-02 manuscript/coexistence-stability_ms.tex
adds PBA final edits

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to accurately represent real-world dynamics is debated
(Eisenhauer et al. 2016; Wardle 2016). Much of the debate centers around
the fact that BEF experimental protocols do not allow species gains to
offset species losses. Theoretical work on diversity-variability
relationships typically suffers from the same limitation: it recognizes
the role of environmental variability in driving population fluctuations
which destabilize ecosystems, but ignores the potential for
environmental variability to promote species richness and thereby help
stabilize ecosystems (Loreau 2010, but see Chesson et al. 2001).

Fluctuating environmental conditions are an important ingredient for
stable species coexistence, both in theoretical models (Chesson 2000a;
Chesson et al. 2004) and in natural communities (C{á}ceres 1997;
Descamps-Julien \& Gonzalez 2005; Adler et al. 2006; Angert et al.
2009). Such ``fluctuation-dependent'' coexistence requires that species
have unique environmental responses and that environmental conditions
vary so that each species in a community experiences favorable and
unfavorable conditions, which prevents competitive exclusion (Chesson
2000a). When coexistence is maintained by a fluctuation-dependent
mechanism, an increase in environmental variability might lead to an
increase in species richness and, consequently, a decrease in ecosystem
variability. However, increasing environmental variability may also
increase ecosystem variability by increasing the fluctuations of
individual species, regardless of species richness. These countervailing
effects of environmental variability present an interesting paradox:
while increasing environmental variability should increase ecosystem
variability, that increase might be offset by an increase in richness.
Such a paradox complicates predictions about how ecosystems will respond
to predicted departures from historical ranges of environmental
variability.

The opposing effects of environmental variability on ecosystem
variability might explain the mixed results from empirical studies on
the diversity-variability relationship. Observational tests of the
diversity-variability relationship, which require sampling across
natural diversity gradients, have yielded negative (Hautier et al.
2014), neutral (Valone \& Hoffman 2003; Cusson et al. 2015), and
positive (Sasaki \& Lauenroth 2011) relationships. In a meta-analysis of
diversity-variability relationships based on observational studies in
terrestrial ecosystems, Jiang \& Pu (2009) found no significant evidence
for an effect of species richness on ecosystem variability.
\tc{blue}{This may be because environmental variability varies across natural diversity gradients, affecting both richness and ecosystem variability.}
The idiosyncratic results of these empirical studies contrast with the
consistent conclusions from theoretical work that ignores feedbacks
between variability and richness.

@@

For the storage effect to operate, we need species-specific responses to
environmental variability, density-dependent covariance between
environmental conditions and competition (\emph{EC} covariance), and
subadditive population growth (Chesson 1994, 2000b). Long-term
coexistence is possible when all species can increase when rare. In the
storage effect, rare species increase by escaping the effects of
\emph{EC} covariance. This happens because common species will
experience greater than average competition (\emph{C}) in good
environment (\emph{E}) years because common species cannot avoid
intraspecific competition. Rare species can escape intraspecific
competition and have the potential to increase rapidly in a good
\emph{E} year. \emph{EC} covariance emerges in our model because
dormant-to-live transition rates (\(\gamma\)) are species-specific and
vary through time. In a high \(\gamma\) year for a common species,
resource uptake will be above average, while in a high \(\gamma\) year
for a rare species, resource uptake will be below average, as long as
\(\gamma\)s are not perfectly correlated between species. Subadditive
population growth buffers populations against large population decreases
in unfavorable years. It is included in our model through a dormant
stage with very low death rates, which limits large population declines
in bad \emph{E} years.

@@

\noindent where \(\sigma^2_{E}\) is the variance of the environmental
cue and \(\rho_{i,j}\) is the correlation between species \emph{i}'s and
species \emph{j}'s transition rates. \(\rho\) must be less than 1 for
stable coexistence, and in all simulations constrained all
\(\rho_{i,j}\)'s to be equal. In a two-species community, the inferior
competitor has the strongest potential to persist when \(\rho=-1\)
(perfectly uncorrelated transition rates). However, in a four-species
community the minimum possible correlation among species is -1/3 given
our constraints that all \(\rho\)'s are equal and that
\(\Sigma(\gamma)\) must be positive-definite. We used the \texttt{R}
function \texttt{mvrnorm} to generate sequences of (un)correlated
variates \emph{E} that we converted to germination rates in the 0-1
range: \(\gamma = e^E / \left(1 + e^E \right)\).

@@

In the absence of environmental fluctuations, the outcome of competition
between two species limited by the same resource is determined by the
shape of their resource uptake curves. That is, at constant resource
supply, whichever species has the lowest resource requirement at
equilibrium (\(R^*\)) will exclude all other species (Tilman 1982).
Resource fluctuations create opportunities for species coexistence
because the resource level is then not fixed at the \(R^*\) of the
superior competitor. If the resource uptake curves of each species are
relatively nonlinear, then some species will be able to take advantage
of resource levels that other species cannot (Chesson 1994).

For example, in Fig. 1C we show two species' resource uptake curves that
are relatively nonlinear. Species B has the lowest \(R^*\) and would
competitively exclude species A in the absence of environmental
fluctuations. But, fluctuating resource supplies can benefit species A
because it can take advantage of relatively high resource levels due its
higher saturation point. Stable coexistence is only possible, however,
if each species creates a disadvantage for itself when abundant. This
occurs in our model because when a resource conservative species (e.g.,
species B in Fig. 1C) is abundant, resource levels will remain high for
a longer period of time after a pulse because its draw down of resources
saturates. Likewise, when a resource acquisitive species (e.g., species
A in Fig. 1C) is abundant, it quickly draws down resources to levels
that favor resource conservative species. Such reciprocity ensures
species can increase when rare and stabilizes coexistence (Armstrong \&
McGehee 1980; Chesson 2000a; Chesson et al. 2004).

@@

When we allowed the variance of the environment to determine which of
four initial species coexisted, similar to a community assembly
experiment, we found a positive relationship between richness and
variability (Fig. 2A,C). This was true for the storage effect, where
coexistence is maintained by fluctuating dormant-to-live transition
rates (\(\gamma\)), and for relative nonlinearity, where coexistence is
maintained by annual resource pulses. The relationship is driven by the
fact that increasing environmental variability increases the strength of
both coexistence mechanisms (Fig. S1-2). More variable conditions
promoted species richness, creating a positive relationship between
diversity and variability.

When we performed species removals but held environmental variability at
a level that allows coexistence of all four species, similar to a
biodiversity-ecosystem functioning experiment, we found a negative
diversity-variability relationship (Fig. 2B,D). Scatter around the
relationship was small for the storage effect because all species has
similar temporal variances. Regardless of species identity, the presence
of more species always stabilizes ecosystem functioning through
portfolio effects. In contrast, scatter around the relationship was
larger for relative nonlinearity (Fig. 2D) because species with
different resource uptake curves have different population variances.
Depending on which species were present, two-species communities were
sometimes less variable than three-species communities.

@@

In communities where species coexist via relative nonlinearity, the
extent to which species additions buffer ecosystem stability against
increases in environmental variability depends on the species traits of
immigrating species and the order in which they enter the community.
When additional species, which immigrate from the regional pool, are
less intrinsically stable than the resident species, ecosystem
variability increases at a constant rate even as species are added (Fig.
4A). If more stable species colonize, species additions buffer the
ecosystem from increasing environmental variability (Fig. 4B). When only
the most unstable (resource acquisitive) species is present, ecosystem
variability initially declines as resource variability increases (Fig.
4B).
\tc{blue}{This occurs because community biomass is very low at low average resource levels for this species, which inflates the *CV*.}

@@

Theory developed for biodiversity-ecosystem function experiments
emphasizes that increases in species richness should reduce ecosystem
variability. Consistent with theoretical expectations from models and
with results from biodiversity-ecosystem functioning experiments, in
which species coexistence is maintained by fluctuation-independent
mechanisms, we produced a negative diversity-variability relationship
(Fig. 2B,D) from a model with fluctuation-dependent species coexistence
(also see Chesson et al. 2001). This agreement is encouraging because
species almost certainly coexist by some combination of
fluctuation-independent (e.g., resource partitioning) and
fluctuation-dependent mechanisms. By extending theory to communities
where species richness is explicitly maintained by temporal variability,
we have gained confidence that experimental findings are generalizable
to many communities. In local settings where environmental variability
is relatively homogeneous, reductions in the number of species should
increase the variability of ecosystem functioning, regardless of how
coexistence is maintained.

When we allowed communities to assemble at sites across a gradient of
environmental variability we discovered a positive relationship between
species richness and ecosystem variability (Fig. 2A,C). While surprising
when viewed through the lens of biodiversity-ecosystem functioning
theory and experimental findings, such a relationship is predicted by
theory on coexistence in fluctuating environments. Environmental
variability is a prerequisite for the storage effect and relative
nonlinearity to stabilize coexistence (Chesson 2000a). These mechanisms
can translate increased variability into higher species richness (Fig.
S1-2), but the increase in environmental variability also increases
ecosystem variability.

@@

While our modeling results show that fluctuation-dependent coexistence
can create positive diversity-variability relationships, whether such
trends are detected will depend on the particular traits of the species
in the community and the relative influence of fluctuation-dependent and
fluctuation-independent coexistence mechanisms. Thus, our results may
also help explain observational studies where no relationship between
diversity and variability is detected. For example, Cusson et al. (2015)
found no relationship between species richness and variability of
abundances in several marine macro-benthic ecosystems. Many of their
focal ecosystems were from highly variable intertidal environments. If
coexistence was at least in part determined by environmental
fluctuations, then the confounding effect of variability and species
richness could compensate any direct effect of species richness on
variability. Previous theoretical work showed how environmental
variation can mask the effect of species diversity on ecosystem
productivity when sampling across sites (Loreau 1998). Our mechanistic
model extends that conclusion to ecosystem variability.

Whether coexistence is fluctuation-independent or fluctuation-dependent
becomes especially important when we consider how ecosystem variability
responds to increasing environmental variability. In the
fluctuation-independent case, species richness is essentially fixed
because the species' inequalities that determine coexistence (niche and
fitness differences) are not linked to environmental variability.
Therefore, increasing environmental variability will always increase
ecosystem variability by increasing the fluctuations of individual
species' abundances.

When coexistence is fluctuation-dependent, however, the outcome is less
certain. By simulating communities with different species pool sizes
across a gradient of environmental variability, we showed that species
gains due to increasing environmental variability can buffer the direct
effect of environmental variability on ecosystem variability (Figs. 3
and 4). These results lead to two conclusions.

First, when predicting the impacts of increasing environmental
variability on ecosystem variability, the mechanism of coexistence in
the community matters. Fluctuation-dependent coexistence can buffer
ecosystems from increasing environmental variability by promoting
increased species richness. Whether our theoretical predictions hold in
real communities is unknown and requires empirical tests. Doing so would
require manipulating environmental variability in communities where
coexistence is known to be fluctuation-dependent, at least in part. Such
data do exist (Angert et al. 2009), and a coupled modeling-experimental
approach could determine if our predictions hold true in real
communities.

>>>>

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to accurately represent real-world dynamics is debated
(Eisenhauer et al. 2016; Wardle 2016). Much of the debate centers on the
fact that BEF experimental protocols do not allow species gains to
offset species losses. Theoretical work on diversity-variability
relationships typically suffers from the same limitation: it recognizes
the role of environmental variability in driving population fluctuations
which destabilize ecosystems, but ignores the potential for
environmental variability to promote species richness and thereby help
stabilize ecosystems (Loreau 2010, but see Chesson et al. 2001).

Fluctuating environmental conditions are an important ingredient for
stable species coexistence, both in theoretical models (Chesson 2000a;
Chesson et al. 2004) and in natural communities (C{á}ceres 1997;
Descamps-Julien \& Gonzalez 2005; Adler et al. 2006; Angert et al.
2009). Such ``fluctuation-dependent'' coexistence requires that species
have unique environmental responses and that environmental conditions
vary so that each species in a community experiences favorable and
unfavorable conditions, which prevents competitive exclusion (Chesson
2000a). When coexistence is maintained by a fluctuation-dependent
mechanism, an increase in environmental variability might lead to an
increase in species richness and, consequently, a decrease in ecosystem
variability. However, increasing environmental variability may also
increase ecosystem variability by increasing the fluctuations of
individual species, regardless of species richness. These countervailing
effects of environmental variability present an interesting paradox:
while we should expect an increase in environmental fluctuations to
increase ecosystem variability, this increase might be buffered if
fluctuation-dependent coexistence adds new species to the community.
Such a paradox complicates predictions about how ecosystems will respond
to predicted departures from historical ranges of environmental
variability.

The opposing effects of environmental variability on ecosystem
variability might explain the mixed results from empirical studies on
the diversity-variability relationship. Observational tests of the
diversity-variability relationship, which require sampling across
natural diversity gradients, have yielded negative (Hautier et al.
2014), neutral (Valone \& Hoffman 2003; Cusson et al. 2015), and
positive (Sasaki \& Lauenroth 2011) relationships. In a meta-analysis of
diversity-variability relationships based on observational studies in
terrestrial ecosystems, Jiang \& Pu (2009) found no significant evidence
for an effect of species richness on ecosystem variability, perhaps
because environmental variability varies across natural diversity
gradients, affecting both richness and ecosystem variability. The
idiosyncratic results of these empirical studies contrast with the
consistent conclusions from theoretical work that ignores feedbacks
between variability and richness.

@@

For the storage effect to operate, we need species-specific responses to
environmental variability, density-dependent covariance between
environmental conditions and competition (\emph{EC} covariance), and
subadditive population growth (Chesson 1994, 2000b). Long-term
coexistence is possible when all species can increase when rare. In the
storage effect, rare species increase by escaping the effects of
\emph{EC} covariance. Common species will experience greater than
average competition (\emph{C}) in good environment (\emph{E}) years
because common species cannot avoid intraspecific competition. However,
rare species can escape intraspecific competition and have the potential
to increase rapidly in a good \emph{E} year. \emph{EC} covariance
emerges in our model because dormant-to-live transition rates
(\(\gamma\)) are species-specific and vary through time. In a high
\(\gamma\) year for a common species, resource uptake will be above
average, while in a high \(\gamma\) year for a rare species, resource
uptake will be below average, as long as \(\gamma\)s are not perfectly
correlated between species. Subadditive population growth buffers
populations against large population decreases in unfavorable years. It
is included in our model through a dormant stage with very low death
rates, which limits large population declines in bad \emph{E} years.

@@

\noindent where \(\sigma^2_{E}\) is the variance of the environmental
cue and \(\rho_{i,j}\) is the correlation between species \emph{i}'s and
species \emph{j}'s transition rates. \(\rho\) must be less than 1 for
stable coexistence, and in all simulations we constrained all
\(\rho_{i,j}\)'s to be equal. In a two-species community, the inferior
competitor has the greatest potential to persist when \(\rho=-1\)
(perfectly uncorrelated transition rates). However, in a four-species
community the minimum possible correlation among species is -1/3 given
our constraints that all \(\rho\)'s are equal and that
\(\Sigma(\gamma)\) must be positive-definite. We used the \texttt{R}
function \texttt{mvrnorm} to generate sequences of (un)correlated
variates \emph{E} that we converted to germination rates in the 0-1
range: \(\gamma = e^E / \left(1 + e^E \right)\).

@@

In the absence of environmental fluctuations, the outcome of competition
between two species limited by the same resource is determined by the
shape of their resource uptake curves. That is, at constant resource
supply, whichever species has the lowest resource requirement at
equilibrium (\(R^*\)) will exclude all other species (Tilman 1982).
Resource fluctuations create opportunities for species coexistence
because the resource level will sometimes exceed the \(R^*\) of the
superior competitor. If the resource uptake curves of each species are
relatively nonlinear, then some species will be able to take advantage
of resource levels that other species cannot (Chesson 1994).

For example, in Fig. 1C we show uptake curves of two species with
different degrees of nonlinearity. Species B has the lowest \(R^*\) and
would competitively exclude species A in the absence of environmental
fluctuations. But fluctuating resource supplies can benefit species A
because it can take advantage of relatively high resource levels due its
higher saturation point. Stable coexistence is only possible, however,
if when each species is dominant it improves conditions for its
competitor. This occurs in our model because when a resource
conservative species (e.g., species B in Fig. 1C) is abundant, it will
draw resources down slowly after a pulse, and its competitor can take
advantage of that period of high resource availability. Likewise, when a
resource acquisitive species (e.g., species A in Fig. 1C) is abundant,
after a pulse it quickly draws down resources to levels that favor
resource conservative species. Such reciprocity helps each species to
increase when rare, stabilizing coexistence (Armstrong \& McGehee 1980;
Chesson 2000a; Chesson et al. 2004).

@@

When we allowed the variance of the environment to determine which of
four initial species coexisted, similar to a study across a natural
diversity gradient, we found a positive relationship between richness
and variability (Fig. 2A,C). This was true for the storage effect, where
coexistence is maintained by fluctuating dormant-to-live transition
rates (\(\gamma\)), and for relative nonlinearity, where coexistence is
maintained by annual resource pulses. The relationship is driven by the
fact that increasing environmental variability increases the strength of
both coexistence mechanisms (Fig. S1-2). More variable conditions
promoted species richness, creating a positive relationship between
diversity and variability.

When we performed species removals but held environmental variability at
a level that allows coexistence of all four species, similar to a
biodiversity-ecosystem functioning experiment, we found a negative
diversity-variability relationship (Fig. 2B,D). Scatter around the
relationship was small for the storage effect because all species have
similar temporal variances. Regardless of species identity, the presence
of more species always stabilized ecosystem functioning through
portfolio effects. In contrast, scatter around the relationship was
larger for relative nonlinearity (Fig. 2D) because species with
different resource uptake curves had different population variances.
Depending on which species were present, two-species communities were
sometimes less variable than three-species communities.

@@

In communities where species coexist via relative nonlinearity, the
extent to which species additions buffer ecosystem stability against
increases in environmental variability depends on the species traits of
immigrating species and the order in which they enter the community.
When additional species, which immigrate from the regional pool, are
less intrinsically stable than the resident species, ecosystem
variability increases at a constant rate even as species are added (Fig.
4A). If more stable species colonize, species additions buffer the
ecosystem from increasing environmental variability (Fig. 4B). When only
the most unstable (resource acquisitive) species is present, ecosystem
variability initially declines as resource variability increases (Fig.
4B) because for this species' monoculture biomass is very low at low
average resource levels for this species, inflating the \emph{CV}.

@@

Theory developed for biodiversity-ecosystem function experiments
emphasizes that increases in species richness should reduce ecosystem
variability. Consistent with theoretical expectations from models and
with results from biodiversity-ecosystem functioning experiments, in
which species coexistence is maintained by fluctuation-independent
mechanisms, our model of fluctuation-dependent species coexistence (also
see Chesson et al. 2001) produced a negative diversity-variability
relationship (Fig. 2B,D). This agreement is encouraging because species
almost certainly coexist by some combination of fluctuation-independent
(e.g., resource partitioning) and fluctuation-dependent mechanisms. By
extending theory to communities where species richness is explicitly
maintained by temporal variability, we have gained confidence that
experimental findings are generalizable to many communities. In local
settings where environmental variability is relatively homogeneous,
reductions in the number of species should increase the variability of
ecosystem functioning, regardless of how coexistence is maintained.

When we allowed communities to assemble at sites across a gradient of
environmental variability, we discovered a positive relationship between
species richness and ecosystem variability (Fig. 2A,C). While surprising
when viewed through the lens of biodiversity-ecosystem functioning
theory and experimental findings, such a relationship is predicted by
theory on coexistence in fluctuating environments. Environmental
variability is a prerequisite for the storage effect and relative
nonlinearity to stabilize coexistence (Chesson 2000a). These mechanisms
can translate increased variability into higher species richness (Fig.
S1-2), but the increase in environmental variability also increases
ecosystem variability.

@@

While our modeling results show that fluctuation-dependent coexistence
can create positive diversity-variability relationships, whether such
trends are detected will depend on the particular traits of the species
in the community and the relative influence of fluctuation-dependent and
fluctuation-independent coexistence mechanisms. Thus, our results may
also help explain observational studies where no relationship between
diversity and variability is detected. For example, Cusson et al. (2015)
found no relationship between species richness and variability of
abundances in several marine macro-benthic ecosystems. Many of their
focal ecosystems were from highly variable intertidal environments. If
coexistence was at least in part determined by environmental
fluctuations, then the confounding effect of variability and species
richness could offset or overwhlem any effect of species richness on
variability. Previous theoretical work showed how environmental
variation can mask the effect of species diversity on ecosystem
productivity when sampling across sites (Loreau 1998). Our mechanistic
model extends that conclusion to ecosystem variability.

Whether coexistence is fluctuation-independent or fluctuation-dependent
becomes especially important when we consider how ecosystem variability
responds to increasing environmental variability. In the
fluctuation-independent case, species richness is essentially fixed
because the niche and fitness differences that determine coexistence are
not linked to environmental variability. Therefore, increasing
environmental variability will always increase ecosystem variability by
increasing the fluctuations of individual species' abundances. When
coexistence is fluctuation-dependent, however, the outcome is less
certain. By simulating communities with different species pool sizes
across a gradient of environmental variability, we showed that species
gains due to increasing environmental variability can buffer the direct
effect of environmental variability on ecosystem variability (Figs. 3
and 4). These results lead to two conclusions.

First, when predicting the impacts of increasing environmental
variability on ecosystem variability, the mechanism of coexistence
matters. Fluctuation-dependent coexistence can buffer ecosystems from
increasing environmental variability by promoting increased species
richness. Whether our theoretical predictions hold in real communities
is unknown and requires empirical tests. Doing so would require
manipulating environmental variability in communities where coexistence
is known to be fluctuation-dependent, at least in part. Such data do
exist (Angert et al. 2009), and a coupled modeling-experimental approach
could determine if our predictions hold true in natural communities.
